\paperTitle

### 2014-01-31 doc/icml2014/directed.tex
changes

%Consider the directed grid model shown in \figureref{grid} which
Consider the directed grid model shown in \figureref{approach}(a) which
  % DONE: all the different dependency => too strong
  captures the local dependency structures possible in
  a Bayesian network.
The model has eight observed variables $x^a_1, x^b_1 \cdots, x^a_4, x^b_4$ and four
  hidden variables $h_1, \cdots, h_4$.
The parameters of this model are the conditional probability tables
$\pi \eqdef \Pr(h_1) \in \Re^k, T \eqdef \Pr(h_2 | h_1) = \Pr(h_3 | h_1) \in \Re^{k \times k},
V \eqdef \Pr(h_4 | h_2, h_3) \in \Re^{k \times k \times k}$ and $O \eqdef \Pr(x^a_i | h_i)
=  \Pr(x^b_i | h_i) \in \Re^{d \times k}$. 
Finally, we assume $O$ and $T$ have full column rank.

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$, allowing us to use
  $\TensorFactorize$, as reviewed in \sectionref{setup}, to recover
  $O$.

@@

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
  A hidden clique $\sC$ of $\sG$ is said to possess \textbf{exclusive views property} if for every hidden variable $h_i \in \sC$,
  there is some observed variable, $x_{v}$, which is conditionally
  independent of the rest of the clique given $h_i$.
  and whose conditional
  moment $\mOpp{v}{i} \eqdef \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.
\end{property}

@@

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any hidden clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$ for which this property holds. Let
  $x_{v_j}$ be an exclusive view for $h_{i_j}$ in $\sC$ (not necessarily unique) and $\sV
  = \{x_{v_1}, \cdots, x_{v_m}\}$ be a set of views for the clique $\sC$.
We can write down the moments $\mO_\sV$ as follows:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{\vh \in \sH}
      \Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh \in \sH} \Pr(\Sh{\sC}) 
          \Pr(x_{i_1} | h_{v_1}) \cdots \Pr(x_{i_m} | h_{v_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE A hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

% DONE: transition the content more gradually (checkpoint: what have we done so far, what next)
% DONE: refer to properties by name in addition to by number
We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}). But which cliques have this property?
In the example above, we were able to guarantee \propertyref{exclusive-views}
  by identifying a hidden variable, $h^1$ along with three conditionally
  independent observed variables, $x^a_1, x^b_1, x^a_2$.
  %followed by using \TensorFactorize to learn the conditional moments.

We define such a hidden variable to be a bottleneck:
\begin{definition}[Bottleneck]
  A hidden variable $h$ is said to be a \textbf{bottleneck} if there exists at
    least three conditionally independent observed variables, or views,
    $x_1, x_2, x_3$. 
  Let $\sV_{h}$ denote a set of views for $h$ (not necessarily unique).
\end{definition}

@@

\begin{figure}
  \centering
  \subfigure[Hidden Markov Model] {
    \includegraphics[width=0.45\columnwidth]{figures/hmm.pdf}
    \label{fig:examples-hmm}
  }
%  \subfigure[Directed grid model] {
%    \label{fig:examples-grid}
%    \includegraphics{figures/grid.pdf}
%  }
  \subfigure[Tree model] {
    \includegraphics[width=0.45\columnwidth]{figures/tree.pdf}
    \label{fig:examples-tree}
  }
  \subfigure[Non-example: a noisy-or network] {
    \includegraphics[height=5em]{figures/non-example.pdf}
    \label{fig:examples-noisy-or}
  }
  \caption{}
  \label{fig:examples}
\end{figure}

@@

In the HMM (\figureref{examples-hmm}), the parameters
are $O \eqdef \Pr(x_i|h_i)$  and that $T \eqdef \Pr(h_{i+1} | h_i)$
for all $i$ (i.e. we have parameter sharing).
While the first (and last) hidden variables $h_1, h_T$ in the
  sequence are not bottlenecks, they still have exclusive views ($x_1$ and
  $x_T$ respectively) whose parameters we know because they share
  parameters, $O$.
%We first use the bottleneck $h_2$ with views $x_1, x_2, x_3$ to estimate $O$,
%and then recover $\pi$ from $\{h_1\}$ and recover $T$ from the clique $\{h_{1}, h_{2}\}$.

@@

The simplest example of a model not identifiable by our algorithm is
  a mixture model with a single view, i.e. \figureref{three-view} with
  just the variables $h_1$ and $x_1$.
However, without further assumptions the model itself is not
  identifiable\footnote{To see this, observe that the probability-mass
  can be arbitrarily exchanged between $\Pr(h_1)$ and $\Pr(x_1 | h_1)$}.
A class of binary-valued noisy-or networks which do not possess the
  bottleneck property are be shown to be identifiable by
  \citet{halpern13noisyor}; 
\figureref{examples-noisy-or} presents one example in this class.

@@

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full
  column rank and that $\pi^{(i)} \eqdef \Pr(h_i) \succ 0$.
But $\mOpp{v}{i}$ is a product of
tensors on the path from $h_i$ to $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following property:%\verify
\begin{assumption} 
\label{asm:full-rank-plus}
For every clique $\sC \in \sG$ (including ones involving observed variables),
every slice of the marginals $Z_\sC$ has full column rank $k$.
\end{assumption}
In our directed grid example (\sectionref{directedExample}), for the clique $\sC = \{ h_1, h_2 \}$ this
  condition implies $\pi = \bP(h_1) \succ 0$ and $T = \bP(h_2 \mid h_1)$ has column rank $k$; 
for the clique $\sC = \{ h_1, x_1 \}$, it implies $O = \bP(x_1 \mid h_1)$ has column rank $k$

@@

From \cite{anandkumar13tensor} given \assumptionref{full-rank-plus},
we have that $\mOpphat{v}{i}$ converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant that
depends polynomially on the $k$-th singular value of $\mOpp{v}{i}$.
Of course, the singular value of $\mOpp{v}{i}$ become exponentially worse as $x_v$ and $h_i$ become farther.
So while we can obtain consistent estimates, the sample complexity
is exponential in the mimimum distance from $h_i$ to any three views, as to be expected.
%(\todo{say something
%quantitative?  annoying part is that we would have to bring in the bounds, so leave it out probably}).

% DONE: need to setup the asymptotic - it's not fresh in everyone's mind.
% Easy case
Next, we need to show that the hidden marginals $\hat Z_\sC$ converge to $Z_\sC$.
First, $\hat M_\sV$ is just an empirical average of multinomials over the data points.
Abusing notation slightly, we let $M_\sV$ also denote its $d^{|\sV|}$-dimensional vectorized version.
We also represent $\mH_\sC$ as
  a vector in $\Re^{k^m}$, and represent $\mOppAll \eqdef
  \mOpp{v_1}{i_1} [x_{v_1}] \otimes \cdots \otimes
  \mOpp{v_m}{i_m} [x_{v_m}]$ as a matrix in $\Re^{d^m \times
  k^m}$.
By the central limit theorem, we have:
$\sqrt{n} (\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV)$,
where $\Sigma_\sV$ is the \emph{asymptotic variance} of $\hat M_\sV$: 
% ARUN: I don't think so - \todo{doesn't this need to be multiplied by something?}
\begin{align*}
\Sigma_\sV \eqdef \dM_\sV - M_\sV M_\sV^\top, \quad \dM_\sV \eqdef \text{diag}(M_\sV).
\end{align*}

@@

We can now use the delta-method to convert the above result into 
the asymptotic variance for the pseudoinverse version of $\LearnClique$:
\begin{lemma}[Asymptotic variance of pseudoinverse estimator for $Z_\sC$]
  \label{lem:mom-variance}  
  Assume $\hat M_{\sV}$ has asymptotic variance $\Sigma_\sV$ defined above.
  Then the asymptotic variance of $\hat Z_\sC$ is:
  \begin{align*}
    \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit,
  \end{align*}
  where $\pinv{\mOppAll} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes
  \cdots \otimes \pinv{\mOpp{v_m}{i_m}}$, a $d^m \times k^m$ matrix.
\end{lemma}
\begin{proof}
For clique $\sC$, recall we have
  $Z_{\sC} = \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m})$.
%Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
%$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
%matrix,
We can rewrite this representation more compactly as: $Z_{\sC} = \pinv{\mOppAll} \mO_\sV$.
By the delta-method \cite{vaart98asymptotic},
% DONE: need to shorten
%we have that the asymptotic variance of $M_\sV$ is:
%\begin{align*}
%  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
%\end{align*}
%where $\Sigma_\sV$ is the variance of the observations. 
we immediately get:
$\sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \mOppAlli \Sigma_\sV \mOppAllit)$.
\end{proof}

>>>>

%Consider the directed grid model shown in \figureref{grid} which
Consider the directed grid model shown in \figureref{approach}(a).
  % DONE: all the different dependency => too strong
  %captures the local dependency structures possible in
  %a Bayesian network.
The model has eight observed variables $x^a_1, x^b_1 \cdots, x^a_4, x^b_4$ and four
  hidden variables $h_1, \cdots, h_4$.
The parameters of this model are the conditional probability tables
$\pi \eqdef \Pr(h_1) \in \Re^k, T \eqdef \Pr(h_2 | h_1) = \Pr(h_3 | h_1) \in \Re^{k \times k},
V \eqdef \Pr(h_4 | h_2, h_3) \in \Re^{k \times k \times k}$ and $O \eqdef \Pr(x^a_i | h_i)
=  \Pr(x^b_i | h_i) \in \Re^{d \times k}$. 
Assume $O$ and $T$ have full column rank.

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$; we can use
  $\TensorFactorize$ (\sectionref{setup}) to recover $O$.

@@

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
  A hidden clique $\sC$ of $\sG$ is said to possess the \textbf{exclusive views
  property} if for every hidden variable $h_i \in \sC$,
  there is some observed variable, $x_{v}$, which is conditionally
  independent of the rest of the clique given $h_i$,
  and whose conditional
  moment $\mOpp{v}{i} \eqdef \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.
\end{property}

@@

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any hidden clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$ for which this property holds. Let
  $x_{v_j}$ be an exclusive view for $h_{i_j}$ in $\sC$ (not necessarily unique) and $\sV
  = \{x_{v_1}, \cdots, x_{v_m}\}$ be a set of views for the clique $\sC$.
We can write down the moments $\mO_\sV$ as follows:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  %&= \sum_{\vh \in \sH}
      %\Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh \in \sH} \Pr(\Sh{\sC}) 
          \Pr(x_{i_1} | h_{v_1}) \cdots \Pr(x_{i_m} | h_{v_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

% DONE: transition the content more gradually (checkpoint: what have we done so far, what next)
% DONE: refer to properties by name in addition to by number
We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}). But in a general graph, which cliques have this property?
In the example above, we were able to guarantee \propertyref{exclusive-views}
  by identifying a hidden variable, $h_1$ along with three conditionally
  independent observed variables, $x^a_1, x^b_1, x^a_2$.
  %followed by using \TensorFactorize to learn the conditional moments.
We define such a hidden variable to be a bottleneck:
\begin{definition}[Bottleneck]
  A hidden variable $h$ is said to be a \textbf{bottleneck} if there exists at
    least three conditionally independent observed variables, or views,
    $x_1, x_2, x_3$. 
  Let $\sV_{h}$ denote a set of views for $h$ (not necessarily unique).
\end{definition}

@@

\begin{figure}
  \centering
  \subfigure[Hidden Markov model] {
    \includegraphics[width=0.45\columnwidth]{figures/hmm.pdf}
    \label{fig:examples-hmm}
  }
%  \subfigure[Directed grid model] {
%    \label{fig:examples-grid}
%    \includegraphics{figures/grid.pdf}
%  }
  \subfigure[Tree model] {
    \includegraphics[width=0.45\columnwidth]{figures/tree.pdf}
    \label{fig:examples-tree}
  }
  \subfigure[Noisy-or model] {
    \includegraphics[height=5em]{figures/non-example.pdf}
    \label{fig:examples-noisy-or}
  }
  \caption{(a) and (b): graphs that satisfy the exclusive views property; (c) does not.}
  \label{fig:examples}
\end{figure}

@@

In the HMM (\figureref{examples-hmm}), the parameters
are $O \eqdef \Pr(x_i|h_i)$  and that $T \eqdef \Pr(h_{i+1} | h_i)$
for all $i$. % (i.e. we have parameter sharing).
While the first and last hidden variables $h_1, h_T$ in the
  sequence are not bottlenecks, they still have exclusive views ($x_1$ and
  $x_T$ respectively) whose parameters we know because they share
  parameters, $O$.
%We first use the bottleneck $h_2$ with views $x_1, x_2, x_3$ to estimate $O$,
%and then recover $\pi$ from $\{h_1\}$ and recover $T$ from the clique $\{h_{1}, h_{2}\}$.

@@

The simplest example of a model that cannot be recovered by our algorithm is
  a mixture model with a single view, i.e. \figureref{three-view} with
  just the variables $h_1$ and $x_1$.
However, without further assumptions the model itself is not
  identifiable\footnote{To see this, observe that the probability-mass
  can be arbitrarily exchanged between $\Pr(h_1)$ and $\Pr(x_1 | h_1)$}.
A class of binary-valued noisy-or networks
(see \figureref{examples-noisy-or} for an example)
which do not possess the
  bottleneck property are nonetheless identifiable;
  see \citet{halpern13noisyor} for an algorithm.

@@

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full
  column rank and that $\pi^{(i)} \eqdef \Pr(h_i) \succ 0$.
But $\mOpp{v}{i}$ is a product of
tensors on the path from $h_i$ to $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following property:%\verify
\begin{assumption} 
\label{asm:full-rank-plus}
For every clique $\sC \in \sG$ (including ones involving observed variables),
every $R^{k \times k}$ slice of the marginals $Z_\sC$ has full column rank $k$.
\end{assumption}
In our directed grid example (\sectionref{directedExample}), for the clique $\sC = \{ h_1, h_2 \}$ this
  condition implies $\pi = \Pr(h_1) \succ 0$ and $T = \Pr(h_2 \mid h_1)$ has column rank $k$; 
for the clique $\sC = \{ h_1, x_1 \}$, it implies $O = \Pr(x_1 \mid h_1)$ has column rank $k$

@@

From \citet{anandkumar13tensor} given \assumptionref{full-rank-plus},
we have that $\mOpphat{v}{i}$ converges to $\mOpp{v}{i}$ at a rate of $n^{-\frac12}$ with a constant that
depends polynomially on the $k$-th singular value of $\mOpp{v}{i}$.
%Of course, the singular value of $\mOpp{v}{i}$ become exponentially worse as $x_v$ and $h_i$ become farther.
%So while we can obtain consistent estimates, the sample complexity
%is exponential in the mimimum distance from $h_i$ to any three views, as to be expected.
%(\todo{say something
%quantitative?  annoying part is that we would have to bring in the bounds, so leave it out probably}).
% DONE: need to setup the asymptotic - it's not fresh in everyone's mind.
% Easy case
Next, we need to show that the hidden marginals $\hat Z_\sC$ converge to $Z_\sC$.
First, $\hat M_\sV$ is just an empirical average of multinomials over the data points.
Abusing notation slightly, we let $M_\sV$ also denote its $d^{|\sV|}$-dimensional vectorized version.
We also represent $\mH_\sC$ as
  a vector in $\Re^{k^m}$, and represent $\mOppAll \eqdef
  \mOpp{v_1}{i_1} \otimes \cdots \otimes
  \mOpp{v_m}{i_m}$ as a matrix in $\Re^{d^m \times
  k^m}$.
By the central limit theorem, we have:
$\sqrt{n} (\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV)$,
where $\Sigma_\sV$ is the \emph{asymptotic variance} of $\hat M_\sV$: 
% ARUN: I don't think so - \todo{doesn't this need to be multiplied by something?}
\begin{align*}
\Sigma_\sV \eqdef \dM_\sV - M_\sV M_\sV^\top, \quad \dM_\sV \eqdef \text{diag}(M_\sV).
\end{align*}

@@

We can now use the delta-method to convert the above result into 
the asymptotic variance for the pseudoinverse version of $\LearnClique$:
\begin{lemma}[Asymptotic variance of pseudoinverse estimator for $Z_\sC$]
  \label{lem:mom-variance}  
  Assume $\hat M_{\sV}$ has asymptotic variance $\Sigma_\sV$ defined above.
  Then the asymptotic variance of $\hat Z_\sC$ is:
  \begin{align*}
    \Sigma^{\mom} &= \mOppAlli \Sigma_\sV \mOppAllit.
  \end{align*}
  %where $\pinv{\mOppAll} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes
  %\cdots \otimes \pinv{\mOpp{v_m}{i_m}}$, a $d^m \times k^m$ matrix.
\end{lemma}
\begin{proof}
For clique $\sC$, recall we have
  $Z_{\sC} = \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m})$.
%Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
%$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
%matrix,
We can rewrite this representation more compactly as: $Z_{\sC} = \pinv{\mOppAll} \mO_\sV$.
By the delta-method \cite{vaart98asymptotic},
% DONE: need to shorten
%we have that the asymptotic variance of $M_\sV$ is:
%\begin{align*}
%  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
%\end{align*}
%where $\Sigma_\sV$ is the variance of the observations. 
we immediately get:
$\sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \mOppAlli \Sigma_\sV \mOppAllit)$.
\end{proof}
